---
title: "Markov Chain Monte Carlo algorithm"
author: "Clement KK Tan"
date: "2024-01-06"
output:
  html_document: default
  word_document: default
---

## R Markdown

## Description of Markov Chain Monte Carlo Algorithm 
The purpose of this algorithm is to generate and simulate random numbers generated in the distribution 
theory for the probability density function given. The answers for Parts A and B are given in this 1 page report. 


# Part A
Part A describes and simulates the random numbers generated from the normal distribution. 

Part A covers the generating of the random walk Metropolis algorithm using N = 10000 and s = 1. It uses the generated samples (x1; : : : xN) to construct a histogram and a kernel density plot in the same figure. Note that these provide estimates of f(x).Overlay a graph of f(x) on this figure to visualize the quality of these estimates.

The rest of the codings and results generated are given in Appendix A steps 1 to 6.


# Part B 
Part B focuses on the convergence diagnostics is the so-called bR value. The convergence diagnostics generate more than one sequence of x0; : : : ; xN, potentially using different initial values x0.

The coding generation involves obtaining the Mj as the sample mean of chain j and Vj as the within sample variance of chain j. After that, use Vj to overall within sample variance W and Mj to obtain the overall sample mean M. Later, use Mj and M to derive the random numbers for the sample variance B. Lastly, use B and W to derive the final parts for the bR values. Calculate the bR for the random walk Metropolis algorithm with N = 2000, s = 0:001 and J = 4. Keeping N and J fixed, provide a plot of the values of bR over a grid of s values in the interval between 0:001 and 1.

The full Part B codings and results are listed in the Appendix Part B steps 1 to 5. 

## Appendix 

# A Step 1: Establish Libraries 
```{r}
library(DBI)        # database information 
library(dplyr)      # Data frame manipulation
library(RSQLite)    #RSQLite for possible SQL usage 
library(lubridate)  # Date operations
library(tidyr)      # Data frame transformation
library(ggplot2)    # Plot any diagrams
library(tidyverse)
library("RColorBrewer")


```


# A Step 2 Set up an initial value x0 as well as a positive integer N and a positive real number s. Then initialize a vector to store the samples
```{r}
x0 = 0
N = 10000
s = 1

samples <- numeric(N)
samples[1] <- x0
```


# A Step 3 Define the distribution function Fx
```{r}
Fx <- function(x) {
  return(0.5 * exp(-abs(x)))
}
```


# A Step 4 Random Walk Metropolis algorithm to get ratio r. After that, accept or reject the proposed sample created.
```{r}
for (i in 2:N) {
  # Simulate a random number from the normal distribution
  x_proposed <- samples[i - 1] + rnorm(1, sd = s)

  # Compute the acceptance ratio
  ratio <- Fx(x_proposed) / Fx(samples[i - 1])

  # Generate a random number from the uniform distribution
  u <- runif(1)

  # Accept or reject the proposed sample
  if (u < ratio) {
    samples[i] <- x_proposed
  } else {
    samples[i] <- samples[i - 1]
  }
}

```


# A Step 5 Create histogram and kernel density plot. Then find true distribution using Fx for comparison.
```{r}
# Plot histogram and kernel density plot
par(mfrow = c(1, 1))
hist(samples, breaks = 50, main = "Histogram and KDE with True Function Overlay", col = "purple", xlab = "Sample x", probability = TRUE, ylab="Density")
lines(density(samples), col = "gold", lwd = 5) # Create KDE line 

# Get the true distribution Fx for comparison
curve(Fx, col = "maroon", add = TRUE, lwd = 3)

legend(1, 0.4, legend=c("RWM Samples", "Fx", "KDE"),col=c("purple", "maroon", "gold"), 
       lty=1:1, cex=0.8)
```


# A Step 6 Obtain and indicate sample mean and standard deviation of the sample in Fx function
```{r}
sample_mean <- mean(samples)
sample_sd <- sd(samples)

cat("Sample Mean:", sample_mean, "\n")
cat("Sample Standard Deviation:", sample_sd, "\n")
```


# B Step 1 Declare parameters for N1, s1, J 
```{r}
J = 4   # Number of chains
N1 = 2000  # Length of each chain
s1 = 0.001  # Standard deviation for random normal values
```


# B Step 2 Generate and calculate the initial chain, sample mean, and within sample variance
```{r}
## Generate initial chain
generate_chain <- function(initial_value, N1, s1) {
  cumsum(rnorm(N1, mean = s1)) + initial_value
}

## Get sample mean 
calculate_sample_mean <- function(chain) {
  mean(chain)
}

## Get within sample variance
calculate_within_sample_variance <- function(chain, sample_mean) {
  mean((chain - sample_mean)^2)
}
```


# B Step 3 Initialise array lists 
```{r}
## Initialize arrays to store sample means, within sample variances, and between sample variances
sample_means <- c()
within_sample_variances <- c()
between_sample_variances <- c()

```


# B Step 4 Define and obtain the different sequences according to the above parameters and chains
```{r}
## Generate and analyze multiple chains
for (j in 1:J) {
  # Generate chain with a random initial value
  initial_value <- rnorm(1)
  chain <- generate_chain(initial_value, N1, s1)


# Calculate sample mean Mj and within sample variance Vj
  Mj <- sample_mean <- calculate_sample_mean(chain)
  Vj <- within_sample_variance <- calculate_within_sample_variance(chain, sample_mean)

  # Store Mj and Vj for each chain
  sample_means <- c(sample_means, Mj)
  within_sample_variances <- c(within_sample_variances, Vj)

  # Print results for each chain
  cat(sprintf("Chain %d - Initial Value: %.2f, Sample Mean Mj: %.2f, Within Sample Variance Vj: %.2f\n", j, initial_value, Mj, Vj))
}

# Calculate overall within sample variance (W)
W <- overall_within_sample_variance <- mean(within_sample_variances)

# Print overall within sample variance
cat(sprintf("\nOverall Within Sample Variance (W): %.2f\n", overall_within_sample_variance))

# Calculate overall sample mean (M)
M <- overall_sample_mean <- mean(sample_means)

# Print overall sample mean
cat(sprintf("\nOverall Sample Mean (M): %.2f\n", overall_sample_mean))

# Calculate between sample variance (B)
for (j in 1:J) {
  between_sample_variances <- c(between_sample_variances, (sample_means[j] - overall_sample_mean)^2)
}

# Average the between sample variances B
B <- overall_between_sample_variance <- mean(between_sample_variances)

# Print overall between sample variance B
cat(sprintf("\nOverall Between Sample Variance (B): %.2f\n", overall_between_sample_variance))

# Calculate R value
R <- sqrt((B + W) / W)

cat(sprintf("Convergence diagnostics R Value: %.2f\n", R))
```


# B Step 5 Keeping N1 and J constant, provide a plot of the values of bR over a grid of s values in the interval between 0.001 and 1.
```{r}
# Generate and calculate the initial chain, sample mean, and within sample variance
generate_chain <- function(initial_value, N2, s2) {
  cumsum(rnorm(N2, mean = s2)) + initial_value
}

calculate_sample_mean <- function(chain) {
  mean(chain)
}

calculate_within_sample_variance <- function(chain, sample_mean) {
  mean((chain - sample_mean)^2)
}

calculate_between_sample_variance <- function(sample_means, overall_sample_mean) {
  mean((sample_means - overall_sample_mean)^2)
}

# Use Gelman Rubin Statistics to obtain the plot of the values of bR over a grid of s values in the interval between 0.001 and 1.
gelman_rubin_statistic <- function(sample_means, overall_sample_mean, within_sample_variances) {
  B2 <- calculate_between_sample_variance(sample_means, overall_sample_mean)
  W2 <- mean(within_sample_variances)
  sqrt((B2 + W2) / W2)
}

# Parameters J and N2
J <- 4   # Number of chains
N2 <- 2000  # Length of each chain

# Grid of s2 values
s_values <- seq(0.001, 1, length.out = 100)

# Initialize an array to store bR values
bR_values <- c()

# Iterate s2 over s values
for (s2 in s_values) {
  # Initialize arrays to store sample means and within sample variances
  sample_means <- c()
  within_sample_variances <- c()

  # Generate and analyze multiple chains
  for (j in 1:J) {
    # Generate chain with a random initial value
    initial_value <- rnorm(1)
    chain <- generate_chain(initial_value, N2, s2)

    # Calculate sample mean and within sample variance
    sample_mean <- calculate_sample_mean(chain)
    within_sample_variance <- calculate_within_sample_variance(chain, sample_mean)

    # Store sample mean and within sample variance for each chain
    sample_means <- c(sample_means, sample_mean)
    within_sample_variances <- c(within_sample_variances, within_sample_variance)
  }

  # Calculate overall sample mean (M)
  overall_sample_mean <- mean(sample_means)

  # Calculate bR value
  bR <- gelman_rubin_statistic(sample_means, overall_sample_mean, within_sample_variances)

  # Store and accumulate bR value
  bR_values <- c(bR_values, bR)
}

# Plotting the Gelman Rubin Statistics Graph to show the graph of bR values across the different s values 
plot(s_values, bR_values, type='l', col='cyan', xlab='s values', ylab='bR values', main='Plot of bR over a grid of s values')
```



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
