---
title: "Part 2 Coding for Airports"
author: "Clement KK Tan"
date: "2024-03-27"
output: html_document
---

## R Markdown


# Part 1 Declare library variables, set working directory and create database conn 
The r info for the library packages are shown below. Do set working directory using the function shown below.  
```{r}
library(DBI)        # database information 
library(dplyr)      # Data frame manipulation
library(RSQLite)    #RSQLite for possible SQL usage 
library(lubridate)  # Date operations
library(tidyr)      # Data frame transformation
library(ggplot2)    # Plot any diagrams
library(tidyverse)
library("RColorBrewer")
library(MASS)
library(ISLR)
library(ISLR2)
library(car)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(tinytex)
library(e1071)
library(mlbench)
library(tree)
library(randomForest)
library(bnlearn)
library(rpart)
library(rpart.plot)

# Set the desired working directory
working_directory <- 'C:/Users/Clement Tan/03 codes pack/Practice Assignment 03'
```


## Part 2 ======== write to the database ========
The given R codes are to read in the CSV files from the airports,csv, carriers.csv and plane-data.csv files. 
```{r}
airports <- read.csv("airports.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
planes <- read.csv("plane-data.csv", header = TRUE)
```

## Part 3 ======== Establish Data range =========
The codes are to read in the number of yeas of CSV data files from 1997 to 1998. This is via the variable "Datarange" and it is to connect the CSV files to the conn variable for the SQLite Connection area.
```{r}
library(data.table)

conn <- dbConnect(RSQLite::SQLite(), "airline.db")

for(i in c(1997:1998)) {
 Datarange <- read.csv(paste0(i, ".csv"), header = TRUE)
  if(i == 1997) {
   dbWriteTable(conn, "Datarange", Datarange)
  } else {
   dbWriteTable(conn, "Datarange", Datarange, append = TRUE)
  }
}
```

## Conversion of CRS timings
```{r}
library(dplyr)

#Declare Datarange for CRSArrTime and CRSDepTime
Datarange$CRSArrTime <- ifelse(Datarange$CRSArrTime / 100 > 23.99, 0, Datarange$CRSArrTime)
Datarange$CRSDepTime <- ifelse(Datarange$CRSDepTime / 100 > 23.99, 0, Datarange$CRSDepTime)

# Declare dataframe for CRSArr_Hour and CRSDep_Hour 
Datarange$CRSArr_Hour <- (as.integer(Datarange$CRSArrTime / 100) %% 24)
Datarange$CRSDep_Hour <- (as.integer(Datarange$CRSDepTime / 100) %% 24)
```

## Obtain flight timings
```{r}
# Load necessary libraries
library(dplyr)

# Convert CRSArrTime and CRSDepTime to datetime format and calculate Flight_Time_Mins
Datarange$Flight_Time_Mins <- sapply(1:nrow(Datarange), function(i) {
  dep_time <- as.POSIXct(sprintf("%04d", Datarange[i, "CRSDepTime"]), format = "%H%M", tz = "UTC")
  arr_time <- as.POSIXct(sprintf("%04d", Datarange[i, "CRSArrTime"]), format = "%H%M", tz = "UTC")
  as.numeric(difftime(arr_time, dep_time, units = "mins"))
})
```

## Remove data sets with DepDelay>0  and ArrDelay>0.
```{r}
# Drop rows where Diverted or Cancelled is equal to 0
Datarange_filtered <- Datarange%>% filter(DepDelay > 0 & ArrDelay > 0)


```

## Obtain the 32 variables
```{r}
dim(Datarange_filtered)
max_value <- max(Datarange_filtered$CRSDepTime, na.rm = TRUE)
```

## Get sample values of the datetime formats
```{r}
# Sample 1000 values from CRSDepTime and CRSArrTime columns
sample_indices <- sample(1:nrow(Datarange), 1000)
sampled_CRSDepTime <- Datarange_filtered$CRSDepTime[sample_indices]
sampled_CRSArrTime <- Datarange_filtered$CRSArrTime[sample_indices]

# Loop over the sampled values
for (i in 1:length(sampled_CRSDepTime)) {
  tryCatch({
    # Convert to datetime format and calculate time difference
    dep_time <- as.POSIXct(sprintf("%04d", sampled_CRSDepTime[i]), format = "%H%M", tz = "UTC")
    arr_time <- as.POSIXct(sprintf("%04d", sampled_CRSArrTime[i]), format = "%H%M", tz = "UTC")
   time_diff_seconds <- difftime(arr_time, dep_time, units = "secs")
    print(time_diff_seconds)
  }, error = function(e) {
    # Print error message and values causing the error
    print(paste("Error:", e))
    print(paste("CRSDepTime:", sampled_CRSDepTime[i]))
    print(paste("CRSArrTime:", sampled_CRSArrTime[i]))
  })
}
```

# Filter and display the CRSArrTime and CRSArr_Hour
```{r}
subset(Datarange_filtered, CRSArr_Hour == 24, select = c("CRSDepTime", "CRSDep_Hour"))
Datarange[Datarange_filtered$CRSArr_Hour == 24, c("CRSDepTime", "CRSDep_Hour")]
```

## Obtain the different time delays 
```{r}
# Filter dataframe for DepDelay <= 15 and get the number of rows
min_delay <- nrow(subset(Datarange_filtered, DepDelay <= 15))
print(min_delay)

# Filter dataframe for DepDelay >= 180 and get the number of rows
big_delay <- nrow(subset(Datarange_filtered, DepDelay >= 180))
print(big_delay)

# Filter dataframe for DepDelay between 15 and 180 (exclusive) and get the number of rows
normal_delay <- nrow(subset(Datarange_filtered, DepDelay > 15 & DepDelay < 180))
print(normal_delay)

# Calculate total delay by summing up delays in each category
total_delay <- min_delay + big_delay + normal_delay
print(total_delay)
```

## Draw pie chart for the different delays 
```{r}
# Define the delay values and proportions
delay <- c(979329, 16872, 758472)
prop_delay <- c(0.794524010926829, 0.0058387863622846045, 0.19963720271088642)
labels <- c('min_delay', 'big_delay', 'normal_delay')

# Create a pie chart
pie(delay, labels = labels, main = "Different Departure Delay Types", col = rainbow(length(delay)))

# Add legend
legend("topright", labels, cex = 0.8, fill = rainbow(length(delay)))
```

## Calculate delay types
```{r}
library(dplyr)

# Define functions to calculate delay categories
calculate_big_delay <- function(delay) {
  if (delay >= 180) {
    return(1)  # or any value you want to assign for big_delay
  } else {
    return(0)  # or any value you want to assign for cases where delay < 180
  }
}

calculate_min_delay <- function(delay) {
  if (delay <= 15) {
    return(1)  # or any value you want to assign for min_delay
  } else {
    return(0)  # or any value you want to assign for cases where delay > 15
  }
}

calculate_normal_delay <- function(delay) {
  if (15 < delay & delay < 180) {
    return(1)  # or any value you want to assign for normal_delay
  } else {
    return(0)  # or any value you want to assign for cases outside the condition
  }
}

# Applying the functions to create new columns
Datarange_filtered<- mutate(Datarange_filtered,
                    big_delay = sapply(Datarange_filtered$ArrDelay, calculate_big_delay),
                    min_delay = sapply(Datarange_filtered$ArrDelay, calculate_min_delay),
                    normal_delay = sapply(Datarange_filtered$ArrDelay, calculate_normal_delay))

# Display the updated DataFrame
print(Datarange_filtered$big_delay)
```

# PART A 
When is the best time of day to fly to minimize delays?
```{r}
# Grouping by CRS Departure Hour and summing the delay types
grouped_data_departure <- Datarange_filtered %>%
 group_by(CRSDep_Hour) %>%
  summarise(big_delay = sum(big_delay),
            min_delay = sum(min_delay),
            normal_delay = sum(normal_delay))

# Reshape the data for ggplot
grouped_data_departure <- gather(grouped_data_departure, key = "delay_type", value = "count", -CRSDep_Hour)

# Create the stacked bar plot
ggplot(grouped_data_departure, aes(x = factor(CRSDep_Hour), y = count, fill = delay_type)) +
 geom_bar(stat = "identity") +
  labs(x = "CRS Departure Hour", y = "Number of Flights", title = "Stacked Bar Plot of Departure Delays by CRS Departure Hour from 1997-1998") +
  scale_x_discrete(labels = function(x) as.character(x)) +
  scale_fill_manual(values = c("big_delay" = "blue", "min_delay" = "orange", "normal_delay" = "green")) +
  theme_minimal()
## This code will create a stacked bar plot showing the number of flights with different delay types (big, min, and normal delays) by CRS departure hour. Adjust the colors and themes as needed.
```

# Part A
Best day of week to fly 
```{r}
library(ggplot2)
    
Datarange_filtered %>%
 mutate(DayOfWeek = factor(x=DayOfWeek, levels=c(1,2,3,4,5,6,7), labels=c("1","2","3","4","5","6","7"))) %>%
  ggplot(aes(x=DayOfWeek, y=min_delay, fill=DayOfWeek)) + 
    geom_boxplot(na.rm=TRUE) + 
    ggtitle("Best Day of week to fly") + 
    xlab("Day of the Week (1 = Monday, 7 = Sunday)") + 
    theme(legend.position="none")  
```

# Part B
Do older planes suffer on a year to year basis?

## Create plane info data using the year csv files 
```{r}
Datarange_filtered <- read.csv(file.path(working_directory, '1997.csv'))
plane_data <- read.csv(file.path(working_directory, 'plane-data.csv'))
plane_data <- rename(plane_data, TailNum = tailnum)
head(plane_data)

```

## Merge data sets
```{r}
flight_data <- merge(Datarange_filtered, plane_data, by = "TailNum")
head(flight_data)
```

## Assuming you have combined the 'Year' and 'year' columns into flight_data
```{r}
# Convert the 'year' column to numeric
flight_data <- flight_data %>%
  mutate(year = as.numeric(year))

# Drop rows where 'year' is NA
flight_data <- flight_data %>%
  filter(!is.na(year))

# Perform the subtraction
flight_data <- flight_data %>%
  mutate(Age_Flight = Year - year)

# Check the result
print(flight_data)
```

## Create Histogram for the age of planes range and the arrrival delays range 
```{r}
# Create the joint plot
h <- ggplot(flight_data, aes(x = Age_Flight, y = DepDelay)) +
  geom_point() +
  geom_smooth(method = "lm", color = "purple") +
  labs(x = "Age_Flight", y = "DepDelay") +
  theme_minimal()

# Print the plot
print(h)
```

## Create calculation functions for the different delays 
```{r}
# Function to calculate 'big_delay'
calculate_big_delay <- function(x) {
  ifelse(x$DepDelay >= 180, 1, 0)
}

# Function to calculate 'min_delay'
calculate_min_delay <- function(x) {
  ifelse(x$DepDelay <= 15, 1, 0)
}

# Function to calculate 'normal_delay'
calculate_normal_delay <- function(x) {
  ifelse(x$DepDelay > 15 & x$DepDelay < 180, 1, 0)
}

```

## Create stack bar plots for the flight age categories and respective delays
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Assuming flight_data is your data frame containing flight information

# Define bins for different ranges of flight times (adjust as needed)
bins <- c(0, 6, 12, 18, 24, 30, 36, Inf)

# Create a new column 'Age_Flight_Category' based on the bins
flight_data$Age_Flight_Category <- cut(flight_data$Age_Flight, breaks = bins, labels = FALSE, right = FALSE)

# Grouping by Flight Time Category and summing the delay types
grouped_data_age_flight <- flight_data %>%
 group_by(Age_Flight_Category) %>%
   summarise(
    big_delay = sum(big_delay),
    min_delay = sum(min_delay),
    normal_delay = sum(normal_delay)
  )

# Stacked bar plot using ggplot2
ggplot(grouped_data_age_flight, aes(x = Age_Flight_Category)) +
  geom_bar(aes(y = big_delay), stat = "identity", fill = "red", alpha = 0.7) +
  geom_bar(aes(y = min_delay), stat = "identity", fill = "blue", alpha = 0.7) +
  geom_bar(aes(y = normal_delay), stat = "identity", fill = "green", alpha = 0.7) +
  
  # Adding labels and title
  labs(x = "Age of Flight Category",
       y = "Number of Flights",
       title = "Stacked Bar Plot of Departure Delays by Age of Flight Category in 1997") +
  
  # Set x-ticks to include all flight time categories
  scale_x_continuous(breaks = 1:length(unique(grouped_data_age_flight$Age_Flight_Category)),
                     labels = c(paste0(bins[1:(length(bins)-1)], "-", bins[2:(length(bins)-1)]), 
                                paste0("More than ", bins[length(bins)-1]))) +
  
  # Adding legend
  scale_fill_manual(values = c("red", "blue", "green"),
                    labels = c("> 180 mins delay", "< 15 mins delay", "normal delay")) +
  
  # Rotate x-axis labels if needed
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```


# Part C 
For each year, fit a logistic regression model for the probability of diverted US fights using as many features as possible from attributes of the departure date, the scheduled departure and arrival times, the coordinates and distance between departure and planned arrival airports, and the carrier. Visualize the coefficients across the years.


## Delcare libraries for logistic regression machine learning 
```{r}
# Declare libraries for logistic regression machine learning
library(tidyverse)
library(caret)
library(Metrics)
library(recipes)
library(caret)
library(mlbench)
library(datasets)
library(dplyr)
library(ggplot2)
library(lubridate)
library(pROC)
library(biglm)
library(readr)
library(broom)
```

# Set seed for reproducibility
```{r}
set.seed(123)
```

# Assuming 'Datarange' is your dataframe containing flight data
```{r}
# Pre-process the data to include only necessary columns
Datarange <- Datarange %>%
  mutate(
    Year = as.numeric(Year),
    Month = as.factor(Month),
    DayOfWeek = as.factor(DayOfWeek),
    UniqueCarrier = as.factor(UniqueCarrier),
    Distance = as.numeric(Distance),
    CRSDepTime = as.numeric(CRSDepTime),  # Already in minutes, ensure numeric
    CRSArrTime = as.numeric(CRSArrTime),  # Convert CRSArrTime to numeric if not already
    Diverted = as.factor(Diverted)  # Ensure 'Diverted' is treated as a factor
  ) %>%
  filter(!is.na(Diverted) & Diverted %in% c("0", "1") & !is.na(CRSDepTime) & !is.na(CRSArrTime))
```

# Sample a smaller subset for modeling
```{r}
Sampled_data <- sample_n(Datarange_filtered, size = 10000)
```

# Fit the logistic regression model using the correct variables
```{r}
Logistic_Model <- glm(Diverted ~ Year + Month + DayOfWeek + UniqueCarrier + Distance +
                        CRSDepTime + CRSArrTime,
                      data = Sampled_data,
                      family = binomial(link = "logit"))

# Print the model summary to inspect the coefficients used
summary(Logistic_Model)
```

# Visualize coefficients from the model
```{r}
summary(Logistic_Model)

ggplot(Logistic_Model$coefficients, aes(x = Logistic_Model$Estimate, y = reorder(term, Logistic_Model$Estimate))) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.7, data = Datarange_filtered) +  # Use geom_bar to create bar chart
  labs(x = "Coefficient Estimate", y = "Predictor", title = "Logistic Regression Coefficients") +
  theme_minimal()
```

```{r}
# Visualize coefficients from the model
ggplot(coefficients, aes(x = estimate, y = reorder(term, estimate))) +
  geom_bar(stat = "identity", fill = "skyblue", alpha = 0.7) +  # Use geom_bar for bar chart
  labs(x = "Coefficient Estimate", y = "Predictor", title = "Logistic Regression Coefficients") +
  theme_minimal()
```


# Design the ROC curve
```{r}
# Calculate predicted probabilities using the Logistic Regression model
predicted_probs <- predict(Logistic_Model, newdata = Sampled_data, type = "response")

# Calculate and plot the ROC curve to evaluate Logistic Regression model performance
roc_curve <- roc(response = as.numeric(Sampled_data$Diverted) - 1, predictor = predicted_probs)
roc_data <- data.frame(specificity = 1 - roc_curve$specificities, sensitivity = roc_curve$sensitivities)
ggplot(roc_data, aes(x = specificity, y = sensitivity)) +
  geom_line() +
  geom_abline(linetype = "dashed") +
  labs(title = "ROC Curve", x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal()
```


